{\bf {\Large Filtering time series of dissolved oxygen for improved estimates of estuary metabolism}}

### 2014-09-24 abstract.tex
finished revisions from MCM, need to proof

In aquatic ecosystems, time series of \ac{DO} can be used to infer integrated ecosystem processes such as primary production, respiration, and net metabolism.  However, continuous monitoring data at estuaries may reflect variation from both biological and physical processes, such that observed data may produce inaccurate or misleading process estimates.  Statistical techniques that dynamically quantify variation in \ac{DO} and tidal changes over time have the potential to isolate biological signals in \ac{DO} variation.  A weighted regression method was developed to normalize, or detide, the predicted \ac{DO} signal to remove the influence of physical advection on ecosystem metabolism estimates.  The method was tested using a simulation approach to create multiple \ac{DO} time series with known additive components of biological and physical variation on different periods.  The method was further validated using one year of continuous monitoring data at four water quality stations that are part of the \acl{NERRS}. We provide a detailed discussion on use of the method for improving certainty in evaluation of \ac{DO} measurements from sites with strong tidal influences.  Moreover, we propose that the method could expand use of the open-water method for estimating ecosystem metabolism in estuaries given that the approach can produce robust estimates of \ac{DO} that are independent of tidal advection.  In particular, this could facilitate the use of shorter deployment periods for water quality monitors or incomplete time series given that known biases related to water movement could be removed.

>>>>

In aquatic ecosystems, time series of \ac{DO} can be used to infer integrated ecosystem processes such as primary production, respiration, and net metabolism.  However, continuous monitoring data at estuaries may reflect variation from both biological and physical processes, such that observed data may produce inaccurate or misleading process estimates.  Statistical techniques that dynamically quantify variation in \ac{DO} and tidal changes over time have the potential to isolate biological signals in \ac{DO} variation.  A weighted regression method was developed to filter, or detide, the predicted \ac{DO} signal to remove the influence of physical advection on ecosystem metabolism estimates.  The method was tested using a simulation approach to create multiple \ac{DO} time series with known additive components of biological and physical variation on different periods.  The method was further validated using one year of continuous monitoring data at four water quality stations that are part of the \acl{NERRS}. We provide a detailed discussion on use of the method for improving certainty in evaluation of \ac{DO} measurements from sites with strong tidal influences.  Moreover, we propose that the method could expand use of the open-water method for estimating ecosystem metabolism in estuaries given that the approach can produce robust estimates of \ac{DO} that are independent of tidal advection.  In particular, this could facilitate the use of shorter deployment periods for water quality monitors or incomplete time series given that known biases related to water movement could be removed.

### 2014-09-24 assess.Rnw
finished revisions from MCM, need to proof

\ac{NERRS} is a federally-funded network of 28 protected estuaries established for long-term research, water-quality monitoring, education, and coastal stewardship \citep{Wenner04}.  Continuous water quality data have been collected at \ac{NERRS} sites since 1994 through the \ac{SWMP}.  In addition to providing a basis for trend evaluation, data from \ac{SWMP} provides an ideal opportunity to evaluate long-term variation in water quality parameters from biological and physical processes.  Continuous \ac{SWMP} data can be used to describe \ac{DO} variation at sites with different characteristics, including variation from ranges in tidal regime \citep{Sanger02} and rates of ecosystem production \citep{Caffrey03,Caffrey04}.

Continuous \ac{DO} time series and tidal height measurements at four sites from the \ac{SWMP} database \citepalias{CDMO14} were used to validate the detiding model with real data.  Monitoring data from January 1\textsuperscript{st} to December 31\textsuperscript{st} 2012 were obtained from a range of geographic locations (\cref{fig:case_map,tab:case_att}). Astronomical tidal heights were predicted for each site using harmonic regression applied to the sonde depth data (\texttt{oce} package in R, \citealt{Foreman89}, \citetalias{RDCT14}). Although, the depth data represent tidal height variation from both astronomical (i.e., gravitational effects) and meteorological (e.g., wind, precipitation inflows) sources, we isolated the former given that daily metabolism estimates were more likely to be affected by repeated diel cycling from normal tidal changes.  Each station was also chosen based on highly significant correlations between \ac{DO} and tidal changes.  The four sites included the Vierra Mouth station at Elkhorn Slough (California, 36.81$^{\circ}$N, 121.78$^{\circ}$W), Bayview Channel at Padilla Bay (Washington, 48.50$^{\circ}$N 122.50$^{\circ}$W), Middle Blackwater River station at Rookery Bay (Florida, 25.93$^{\circ}$N 81.60$^{\circ}$W), and Dean Creek station at Sapelo Island (Georgia, 31.39$^{\circ}$N 81.28$^{\circ}$W).  The stations are generally macrotidal semidiurnal or mixed semidiurnal and net heterotrophic on an annual basis (\cref{tab:case_att}).  Net heterotrophy (i.e., respiration exceeding production) is typical for shallow water systems at temperate latitudes \citep{Caffrey03}, although values in \cref{tab:case_att} are from observed \ac{DO} time series that are strongly influenced by tidal advection.

\subsection{Estimates of ecosystem metabolism before and after detiding} \label{met_sec}

The weighted regression method was applied to the time series for each station to obtain a detided \ac{DO} time series for estimating metabolism.  Half window widths of six days, one hour, and a tidal proportion of one half were chosen based on a balance between large and small window widths, although the chosen window widths are arbitrary and a more exhaustive evaluation should be conducted prior to using the results to inform management. Unlike the simulated data, the true biological \ac{DO} signal was unknown for the case studies.  Accordingly, results were evaluated  using correlations of \ac{DO} and metabolism estimates with tidal height before and after application of the model.  Daily metabolism estimates before and after detiding were compared to the mean rate of tidal height change (i.e., first derivative of the predicted tidal height) for each day during separate solar periods.  Production rates were compared to mean rates of tidal height change during the day, respiration rates were compared to mean rates of change during the night, and net metabolism rates were compared to mean rates of change for the total 24 hour period each day.  Results were also evaluated based on the occurrence of `anomalous' daily production or respiration estimates, where anomalous was defined as negative production during the day and positive respiration estimates during the night.  Anomalous values have been previously attributed to the effects of physical processes on \ac{DO} time series \citep{Caffrey03}.  We hypothesized that metabolism esimates using the detided signal would contain less `anomalous' values than those from the observed \ac{DO} time series.  Although anomalies could be caused by processes other than tidal advection, e.g., abiotic dark oxygen production \citep{Pamatmat97}, we assume that physical processes are the dominant sources of these values given the tidal characteristics at each site.  Finally, means and standard errors of metabolism estimates were evaluated before and after detiding to determine if annual aggregations were significantly different.

Ecosystem metabolism was estimated from the \ac{DO} time series using the open-water technique \citep{Odum56} as described in \citet{Caffrey13}.  The method is used to infer net ecosystem metabolism using the mass balance equation:
\begin{equation} \label{metrate}
\frac{\delta DO}{\delta t} = P - R + D
\end{equation}
where the change in \ac{DO} concentration ($\delta DO$, g O$_2$ m$^{-3}$) over time ($\delta t$, hours) is equal to photosynthetic rate ($P$, g O$_2$ m$^{-3}$ hr$^{-1}$) minus respiration rate ($R$, g O$_2$ m$^{-3}$ hr$^{-1}$), corrected for air-sea gas exchange at the surface ($D$, g O$_2$ m$^{-3}$ hr$^{-1}$) \citep{Caffrey13}. $D$ is estimated as the difference between the \ac{DO} saturation concentration and observed \ac{DO}, multiplied by a volumetric reaeration coefficient, $k_a$ \citep{Thebault08}.  The diffusion-corrected \ac{DO} flux estimates were averaged during day and night for each 24 hour period in the time series, where flux is an hourly rate of \ac{DO} change as the difference between observations at time $t$ and $t+1$ on an hourly scale.  Areal respiration rates were assumed constant during the night and substracted from daily gross production estimates to yield net ecosystem metabolism (\cref{tab:case_att}).

Detiding had significant effects on the correlations between tidal height changes, \ac{DO} time series, and metabolism estimates (\cref{tab:cor_res},  see the link in the \hyperref[multi]{multimedia} section for graphical results of each case study).  Correlations of observed \ac{DO} time series with predicted tidal height were highly significant, with all sites indicating positive relationships, except Padilla Bay where tidal increases were associated with declines in \ac{DO}  concentration.  This suggests that seaward water masses were less anoxic than landward masses, with the opposite being true for Padilla Bay. The detided \ac{DO} time series had greatly reduced correlations with tidal height, although relationships were still significant after detiding likely because of the large sample size for each site (n $\approx$ 17500). Comparison of metabolic rates to tidal changes before and after detiding produced inconsistent results (\cref{tab:cor_res}).  Correlations for Elkhorn Slough and Sapelo Island showed consistent reductions in all three metabolims estimates after detiding.  Correlations for Padilla Bay and Rookery Bay were of opposite sign and greater magnitude after detiding for production and respiration, although net metabolism estimates had reduced correlations.

The proportion of daily integrated metabolism estimates that were anomalous (negative production, positive respiration) were significantly reduced for most sites after detiding (\cref{tab:case_res}).  Before detiding, anomalous values ranged from \Sexpr{form_fun(min(anom_obs[anom_obs$Metab == 'Pg', 'Value']))} (Rookery Bay) to \Sexpr{form_fun(max(anom_obs[anom_obs$Metab == 'Pg', 'Value']))} (Padilla Bay) for production and \Sexpr{form_fun(min(anom_obs[anom_obs$Metab == 'Rt', 'Value']))} (Rookery Bay) to \Sexpr{form_fun(max(anom_obs[anom_obs$Metab == 'Rt', 'Value']))} (Elkhorn Slough) for respiration. Anomalous values were reduced to near zero for Rookery Bay and Sapelo Island, by approximately half for Padilla Bay (\Sexpr{form_fun(anom_dtd[anom_dtd$Metab == 'Pg' & anom_dtd$.id == 'Padilla Bay', 'Value'])} for production, \Sexpr{form_fun(anom_dtd[anom_dtd$Metab == 'Rt' & anom_dtd$.id == 'Padilla Bay', 'Value'])} for respiration), and only slightly reduced for Elkhorn Slough (\Sexpr{form_fun(anom_dtd[anom_dtd$Metab == 'Pg' & anom_dtd$.id == 'Elkhorn Slough', 'Value'])} for production, \Sexpr{form_fun(anom_dtd[anom_dtd$Metab == 'Rt' & anom_dtd$.id == 'Elkhorn Slough', 'Value'])} for respiration).  Metabolism estimates using detided \ac{DO} time series had decreased mean production and respiration for Elkhorn Slough, increased mean production and respiration for Padilla Bay, and generally unchanged mean production and respiration for Rookery Bay and Sapelo Island (\cref{tab:case_res}).  Mean net ecosystem metabolism was unchanged for all sites.  Decreases in the standard erorr for all metabolism estimates (production, respiration, and net) were observed for all cases after detiding.

An example from Sapelo Island illustrates the effects of weighted regression on \ac{DO} and metabolism estimates (\cref{fig:phase_out,fig:phase_in,fig:case_ex}).   A two-week period in February shows when the tidal changes were both in and out of phase with the diel cycling.  Maximum tide heights were generally out of phase with the diel cycle during the first week such that low tides were observed during the middle of the night and the middle of the day (\cref{fig:phase_in}).  The opposite scenario occurred during the second week when the maximum tide height occured during the day and night (\cref{fig:phase_in}).  The effects of tidal height change on the observed \ac{DO} time series are visually apparent in the plots. The first week illustrates a strong negative bias (less respiration, less production) in the observed \ac{DO} signal from low tides at mid-day and mid-night, whereas the second example illustrates a strong positive bias (more respiration, more production) in the observed \ac{DO} from high tides. These biases are apparent in the metabolism estimates using the observed data (\cref{fig:case_ex}).  Anomalous estimates occur when low tides are in phase with the solar cycle, whereas metabolism estimates are likely over-estimated when high tides are in phase with the solar cycle.  The detided time series shows noticeable changes given the direction of bias from the phasing between tidal height and diel period.  \ac{DO} values were higher after detiding when low tides occurred during night and day periods, whereas \ac{DO} values were lower after detiding when high tides occurred during day and night periods (\cref{fig:phase_out,fig:phase_in}).  Changes in metabolism estimates after detiding were also apparent, such that the anomalous values were removed during the first week and the positive bias in the second week is decreased.  Detiding had similar effects for the remaining sites, particularly when tidal changes were strongly in or out of phase with diel periods.

\subsection{Effects of aggregation and importance of detiding}

A final point of concern is the period of observation within which observed \ac{DO} is affected by tidal height changes and the extent to which this affects the interpretation of ecosystem metabolism.  From a management or ecological perspective, the effects of tidal variation on daily estimates may not be a primary concern given that seasonal or annual rates may be more relevant for evaluating ecosystem dynamics with continuous monitoring data.  The example from Sapelo Island in the previous section highlights this point given that mean production and respiration estimates before and after detiding were generally unchanged for the two-week period. \cref{tab:case_res} also indicated that mean annual estimates of production and respiration were unchanged for Rookery Bay and Sapelo Island, whereas production and respiration estimates were significantly different for Elkhorn Slough and Padilla Bay. Although we acknowledge that the specific results may be related to the window widths, this suggests that tidal effects contribute to changes in metabolic estimates from observed time series that are aggregated on longer time periods.  Therefore, an evaluation of the effects of tidal variation on ecosystem metabolism for different periods of observation is critical for understanding practical implications of weighted regression.  Specifically, when should detiding be applied if aggregation of observed data on longer time periods removes potential bias?  A comparison of observed and detided estimates that are aggregated over different periods of observation (e.g., annual, seasonal, monthly) could help address this question.

The observed and detided daily estimates were averaged by month and season (Fall, Spring, Summer, and Winter) for each case study to evaluate effects of aggregation on mean production and respiration estimates.  Mean annual estimates in \cref{tab:case_res} also provided a basis of comparison with monthly and seasonal aggregation. Significant variation in aggregated production and respiration estimates for month and season was observed for each case study (\cref{fig:metab_sum1,fig:metab_sum2}).  Detided production and respiration estimates for Padilla Bay and Rookery Bay exhibited seasonal and monthly variation that was more characteristic of expected trends with increases in metabolism during warmer months.  Specifically, production estimates based on observed \ac{DO} were substantially muted for both Padilla Bay (\cref{fig:metab_sum1}) and Rookery Bay (\cref{fig:metab_sum2}) during summer months, whereas values were significantly higher based on the detided data. Results for Sapelo Island suggested that winter and summer months were under- and over-estimated, respectively, based on the observed data.  Results for Elkhorn Slough varied significantly such that production and respiration were significantly reduced after detiding regardless of the aggregation period.  Overall, these trends emphasize the importance of considering different aggregation periods for interpreting metabolism estimates.  Each case study showed differences in observed and detided values at monthly and seasonal aggregations, whereas only two of the four case studies had mean aggregated estimates that were substantially different (Elkhorn Slough and Padilla Bay, \cref{tab:case_res}).  Periods of observation as long as one year may include significant sources of bias from tidal advection, suggesting the need for applying weighted regression given careful consideration of appropriate window widths.

>>>>

Continuous monitoring data from the \acl{NERRS} was used to validate the weighted regression model by evaluating estimates of ecosytem metabolism obtained from observed and filtered \ac{DO} time series. \ac{NERRS} is a federally-funded network of 28 protected estuaries established for long-term research, water-quality monitoring, education, and coastal stewardship \citep{Wenner04}.  Continuous water quality data have been collected at \ac{NERRS} sites since 1994 through the \aclu{SWMP} (\acs{SWMP}, \citetalias{CDMO14}).  In addition to providing a basis for trend evaluation, data from \ac{SWMP} provides an ideal opportunity to evaluate long-term variation in water quality parameters from biological and physical processes.  Continuous \ac{SWMP} data can be used to describe \ac{DO} variation at sites with different characteristics, including variation from ranges in tidal regime \citep{Sanger02} and rates of ecosystem production \citep{Caffrey03,Caffrey04}.  For the purpose of this analysis, we selected sites from the \ac{SWMP} database that had desirable characteristics for validating weighted regression.  Specifically, four macrotidal sites were chosen based on apparent relationships between \ac{DO} and tidal changes (\cref{fig:case_map,tab:case_att}): Vierra Mouth station at Elkhorn Slough (California, 36.81$^{\circ}$N, 121.78$^{\circ}$W), Bayview Channel at Padilla Bay (Washington, 48.50$^{\circ}$N 122.50$^{\circ}$W), Middle Blackwater River station at Rookery Bay (Florida, 25.93$^{\circ}$N 81.60$^{\circ}$W), and Dean Creek station at Sapelo Island (Georgia, 31.39$^{\circ}$N 81.28$^{\circ}$W).

We analyzed continuous \ac{DO} time series and water level measurements from January 1\textsuperscript{st} to December 31\textsuperscript{st} 2012 at the four sites to evaluate the regression model.  Tide predictions were obtained for each site using harmonic regression applied to the sonde depth data (\texttt{oce} package in R, \citealt{Foreman89}, \citetalias{RDCT14}). The stations were generally semidiurnal or mixed semidiurnal and net heterotrophic on an annual basis (\cref{tab:case_att}).  Net heterotrophy (i.e., respiration exceeding production) is typical for shallow water systems at temperate latitudes \citep{Caffrey03}, although values in \cref{tab:case_att} were from observed \ac{DO} time series that were strongly correlated with water level height.

\subsection{Estimates of ecosystem metabolism before and after filtering} \label{met_sec}

The weighted regression method was applied to the time series for each station to obtain a filtered \ac{DO} time series for estimating metabolism.  Ecosystem metabolism was estimated from the \ac{DO} time series using the open-water technique \citep{Odum56} as described in \citet{Caffrey13}.  The method is used to infer net ecosystem metabolism using the mass balance equation:
\begin{equation} \label{metrate}
\frac{\delta DO}{\delta t} = P - R + D
\end{equation}
where the change in \ac{DO} concentration ($\delta DO$, g O$_2$ m$^{-3}$) over time ($\delta t$, hours) is equal to photosynthetic rate ($P$, g O$_2$ m$^{-3}$ hr$^{-1}$) minus respiration rate ($R$, g O$_2$ m$^{-3}$ hr$^{-1}$), corrected for air-sea gas exchange ($D$, g O$_2$ m$^{-3}$ hr$^{-1}$) \citep{Caffrey13}. $D$ is estimated as the difference between the \ac{DO} saturation concentration and observed \ac{DO} concentration, multiplied by a volumetric reaeration coefficient, $k_a$ \citep{Thebault08}.  The diffusion-corrected \ac{DO} flux estimates were averaged during day and night for each 24 hour period in the time series, where flux is an hourly rate of \ac{DO} change.  Rspiration rates were assumed constant during the night and substracted from daily net production estimates to yield gross production (\cref{tab:case_att}).

Half window widths of six days, one hour, and a tidal proportion of one half were used to filter the observed \ac{DO} time series.  Unlike the simulated data, the true biological \ac{DO} signal was unknown for the case studies.  Accordingly, the regression results were evaluated  using correlations of \ac{DO} and metabolism estimates with tidal height before and after application of the model.  Daily metabolism estimates before and after filtering were compared to the mean rate of tidal height change (i.e., first derivative of the predicted tidal height) for each day during separate solar periods.  Production rates were compared to mean rates of tidal height change during the day, respiration rates were compared to mean rates of change during the night, and net metabolism rates were compared to mean rates of change for the total 24 hour period each day.  Results were also evaluated based on the occurrence of `anomalous' daily production or respiration estimates, where anomalous was defined as negative production during the day and positive respiration estimates during the night.  Anomalous values have been previously attributed to the effects of physical processes on \ac{DO} time series \citep{Caffrey03}.  We hypothesized that metabolism esimates using the filtered signal would contain less `anomalous' values than those from the observed \ac{DO} time series.  Although anomalies could be caused by processes other than tidal advection, e.g., abiotic dark oxygen production \citep{Pamatmat97}, we assume that physical processes are the dominant sources of these values given the tidal characteristics at each site.  Finally, means and standard errors of metabolism estimates were evaluated before and after filtering to determine if annual aggregations were significantly different.

Filtering had significant effects on the correlations between water level changes, \ac{DO} time series, and daily integrated metabolism estimates (\cref{tab:cor_res}, see the link in the \hyperref[multi]{multimedia} section for graphical results of each case study).  Correlations of observed \ac{DO} time series with predicted tidal height were highly significant and positive at all sites, except Padilla Bay where increases in water level were associated with decreases in \ac{DO}  concentration.  The filtered \ac{DO} time series had greatly reduced correlations with tidal height, although relationships were still significant after filtering likely because of the large sample size for each site (n $\approx$ 17,500). Comparison of metabolic rates to tidal changes before and after filtering produced inconsistent results (\cref{tab:cor_res}).  Correlations for Elkhorn Slough and Sapelo Island showed consistent reductions in all three metabolims estimates after filtering.  Correlations for Padilla Bay and Rookery Bay were of opposite sign and greater magnitude after filtering for production and respiration, although net metabolism estimates had reduced correlations.

The proportion of daily integrated metabolism estimates that were anomalous (negative production, positive respiration) were significantly reduced for most sites after filtering (\cref{tab:case_res}), perhaps indicating the relative effects of water movement.  Before filtering, anomalous values ranged from \Sexpr{form_fun(min(anom_obs[anom_obs$Metab == 'Pg', 'Value']))} (Rookery Bay) to \Sexpr{form_fun(max(anom_obs[anom_obs$Metab == 'Pg', 'Value']))} (Padilla Bay) for production and \Sexpr{form_fun(min(anom_obs[anom_obs$Metab == 'Rt', 'Value']))} (Rookery Bay) to \Sexpr{form_fun(max(anom_obs[anom_obs$Metab == 'Rt', 'Value']))} (Elkhorn Slough) for respiration. Anomalous values were reduced to near zero for Rookery Bay and Sapelo Island, by approximately half for Padilla Bay (\Sexpr{form_fun(anom_dtd[anom_dtd$Metab == 'Pg' & anom_dtd$.id == 'Padilla Bay', 'Value'])} for production, \Sexpr{form_fun(anom_dtd[anom_dtd$Metab == 'Rt' & anom_dtd$.id == 'Padilla Bay', 'Value'])} for respiration), and only slightly reduced for Elkhorn Slough (\Sexpr{form_fun(anom_dtd[anom_dtd$Metab == 'Pg' & anom_dtd$.id == 'Elkhorn Slough', 'Value'])} for production, \Sexpr{form_fun(anom_dtd[anom_dtd$Metab == 'Rt' & anom_dtd$.id == 'Elkhorn Slough', 'Value'])} for respiration).  Metabolism estimates using filtered \ac{DO} time series had decreased mean production (\Sexpr{form_fun(elk_per[, 'Pg_Avg'],1,1,1)} \%) and respiration (\Sexpr{form_fun(elk_per[, 'Rt_Avg'],1,1,1)} \%) for Elkhorn Slough, increased mean production (\Sexpr{form_fun(pdb_per[, 'Pg_Avg'],1,1,1)} \%) and respiration (\Sexpr{form_fun(pdb_per[, 'Rt_Avg'],1,1,1)} \%) for Padilla Bay, and generally unchanged mean production and respiration for Rookery Bay and Sapelo Island (\cref{tab:case_res}).  Mean net ecosystem metabolism was unchanged for all sites.  Decreases in the standard erorr for all metabolism estimates (production, respiration, and net) were observed for all cases after filtering.

An example from Sapelo Island illustrates the effects of weighted regression on \ac{DO} and metabolism estimates (\cref{fig:phase_out,fig:phase_in,fig:case_ex}).   A two-week period in February showed when the tidal cycles were both in and out of phase with the diel cycling, where phasing describes synchronicity between maximum tide heights and day/night periods.  That is, maximum tide heights were generally out of phase with the diel cycle during the first week when low tides were observed during the middle of the night and the middle of the day (\cref{fig:phase_in}), whereas tide heights were in phase during the second week when the maximum tide height occured during the day and night (\cref{fig:phase_in}).  The effects of tidal height change on the observed \ac{DO} time series are visually apparent in the plots. The first week illustrates a strong negative bias (less respiration, less production) in the observed \ac{DO} signal from low tides at mid-day and mid-night, whereas the second example illustrates a strong positive bias (more respiration, more production) in the observed \ac{DO} from high tides. These biases are apparent in the metabolism estimates using the observed data (\cref{fig:case_ex}).  Anomalous estimates occur when low tides are in phase with the solar cycle, whereas metabolism estimates are likely over-estimated when high tides are in phase with the solar cycle.  The filtered time series shows noticeable changes given the direction of bias from the phasing between tidal height and diel period.  \ac{DO} values were higher after filtering when low tides occurred during night and day periods, whereas \ac{DO} values were lower after filtering when high tides occurred during day and night periods (\cref{fig:phase_out,fig:phase_in}).  Changes in metabolism estimates after filtering were also apparent, such that the anomalous values were removed during the first week and the positive bias in the second week is decreased.  filtering had similar effects for the remaining sites, particularly when tidal changes were strongly in or out of phase with diel periods.

\subsection{Effects of aggregation and importance of filtering}

A point of concern is the period of observation within which observed \ac{DO} is affected by tidal height changes and the extent to which this affects the interpretation of ecosystem metabolism.  The effects of tidal variation on daily estimates may not be relevant if seasonal or annual aggregations remove this potential bias.  The example from Sapelo Island in the previous section highlights this point given that mean production and respiration estimates before and after filtering were generally unchanged for the two-week period. \cref{tab:case_res} also indicated that mean annual estimates of production and respiration were unchanged for Rookery Bay and Sapelo Island.  However, annual averages of production and respiration estimates were significantly different for Elkhorn Slough and Padilla Bay. Given these results, tidal variation may have effects on metabolism estimates on time scales much longer than 24 hours depending on the location.  Therefore, an evaluation of the weighted regression analysis to filter the effects of tidal variation on ecosystem metabolism for different periods of observation is critical for its application.  Specifically, when should filtering be applied if aggregation of observed data on longer time periods removes potential bias?  A comparison of observed and filtered estimates that are aggregated over different periods of observation (e.g., annual, seasonal, monthly) could help address this question.

The observed and filtered daily estimates were averaged by month and season (Fall, Spring, Summer, and Winter) for each case study to evaluate effects of aggregation on mean production and respiration estimates.  Mean annual estimates in \cref{tab:case_res} also provided a basis of comparison with monthly and seasonal aggregation. Significant variation in aggregated production and respiration estimates for month and season was observed for each case study (\cref{fig:metab_sum1,fig:metab_sum2}).  filtered production and respiration estimates for Padilla Bay and Rookery Bay exhibited seasonal and monthly variation that was more characteristic of expected trends with increases in metabolism during warmer months.  Specifically, production estimates based on observed \ac{DO} were substantially muted for both Padilla Bay (\cref{fig:metab_sum1}) and Rookery Bay (\cref{fig:metab_sum2}) during summer months, whereas values were significantly higher based on the filtered data. Results for Sapelo Island suggested that winter and summer months were under- and over-estimated, respectively, based on the observed data.  Results for Elkhorn Slough varied significantly such that production and respiration were significantly reduced after filtering regardless of the aggregation period.  Overall, these trends emphasize the importance of considering different aggregation periods for interpreting metabolism estimates.  Each case study showed differences in observed and filtered values at monthly and seasonal aggregations, whereas only two of the four case studies had mean aggregated estimates that were substantially different (Elkhorn Slough and Padilla Bay, \cref{tab:case_res}).  Periods of observation as long as one year may include significant sources of bias from tidal advection, suggesting the need for applying weighted regression given careful consideration of appropriate window widths.

### 2014-09-24 comments.Rnw
finished revisions from MCM, need to proof

Results from the simulations and case studies suggested that weighted regression can be a practical approach for detiding \ac{DO} time series to remove the effects of physical advection on estimates of ecosystem metabolism.  However, application of the method may only be appropriate under specific situations.  The case studies were chosen based on the relatively high proportion of metabolism estimates that were anomalous and the strength of correlation between the observed \ac{DO} time series and tidal height.  Despite these similarites among the case studies, detiding had variable effects on metabolism estimates.  The results for Elkhorn Slough and Padilla Bay are of particular concern given that mean annual estimates were substantially different compared to those from the observed \ac{DO} time series.  Although the correlation of \ac{DO} and tidal height was reduced for both cases, in addition to a reduction of anomalous estimates, the relative change in mean metabolism before and after detiding suggests a more careful evaluation of the method is needed.  In particular, alternative window widths should evaluated for the ability to remove tidal effects while preserving the biological signal.  The window widths in the above analysis may have removed variation in the \ac{DO} signal from both of these sources.

Although the above analyses suggest the approach has merit, the case studies emphasize a critical challenge in applying weighted regression to monitoring data. Specifically, the true biological signal is not known and the relative contribution of horizontal advection to bias is not accurately quantified with the available data.  Comparative analyses between systems with varying tidal influence or within-system evaluations of multiple sites at fixed distances are necessary to further validate the performance of weighted regression.  In the absence of additional validation, we propose a precautionary approach for application of the weighted regression to monitoring data.  Weighted regression may be most effective at macrotidal sites with strong evidence of the effects of tidal advection on biological signals.  A weight-of-evidence approach should be used such that the occurrence of anomalous metabolism estimates, strong correlations between observed \ac{DO} and tide height, and clear visual patterns of tide change on \ac{DO} would suggest detiding is appropriate.  The choice of window widths may also produce varying results.  Window widths that produce large changes in mean annual estimates should be interpreted with caution.  In general, a pragmatic approach is emphasized such that results should be evaluated based on the preservation of diel variation from production while exhibiting minimal changes with the tide.  Such an approach, combined with further validation, will support informed management decisions through more accurate estimates of ecosystem metabolism.

>>>>

Results from the simulations and case studies suggested that weighted regression can be a practical approach for filtering \ac{DO} time series to remove the effects of physical advection on estimates of ecosystem metabolism.  However, application of the method may only be appropriate under specific situations.  The case studies were chosen based on the relatively high proportion of metabolism estimates that were anomalous and the strength of correlation between the observed \ac{DO} time series and tidal height.  Despite these similarites among the case studies, filtering had variable effects on metabolism estimates.  The results for Elkhorn Slough and Padilla Bay are of particular concern given that mean annual estimates were substantially different compared to those from the observed \ac{DO} time series.  Although the correlation of \ac{DO} and tidal height was reduced for both cases, in addition to a reduction of anomalous estimates, the relative change in mean metabolism before and after filtering suggests a more careful evaluation of the method is needed.  In particular, alternative window widths should evaluated for the ability to remove tidal effects while preserving the biological signal.  The window widths in the above analysis may have removed variation in the \ac{DO} signal from both of these sources.

Although the above analyses suggest the approach has merit, the case studies emphasize a critical challenge in applying weighted regression to monitoring data. Specifically, the true biological signal is not known and the relative contribution of horizontal advection to bias is not accurately quantified with the available data.  Comparative analyses between systems with varying tidal influence or within-system evaluations of multiple sites at fixed distances are necessary to further validate the performance of weighted regression.  In the absence of additional validation, we propose a precautionary approach for application of the weighted regression to monitoring data.  Weighted regression may be most effective at macrotidal sites with strong evidence of the effects of tidal advection on biological signals.  A weight-of-evidence approach should be used such that the occurrence of anomalous metabolism estimates, strong correlations between observed \ac{DO} and tide height, and clear visual patterns of tide change on \ac{DO} would suggest filtering is appropriate.  The choice of window widths may also produce varying results.  Window widths that produce large changes in mean annual estimates should be interpreted with caution.  In general, a pragmatic approach is emphasized such that results should be evaluated based on the preservation of diel variation from production while exhibiting minimal changes with the tide.  Such an approach, combined with further validation, will support informed management decisions through more accurate estimates of ecosystem metabolism.

### 2014-09-24 dep_dat.Rnw
changed means to medians
finished revisions from MCM, need to proof

# used in beginning of assessment
perf_means <- colMeans(mod_perf[, c('cor', 'err')])
perf_rngs <- apply(mod_perf[, c('cor', 'err')], 2, range)

@@

>>>>

# used in beginning of assessment
perf_medians <- colMedians(mod_perf[, c('cor', 'err')])
perf_rngs <- apply(mod_perf[, c('cor', 'err')], 2, range)

@@

######
# changes in mean production, respiration for each site after filtering
# used in estimates of ecosytem metab before/after section in assessment
# basically same code used to create table 5
load('met_comp.RData')

mean_met <- met_comp
names(mean_met)[1] <- 'Site'

# convert metab data to g m^-2 d^-1
#  1mmolO2 = 32 mg O2, 1000mg = 1g, multiply by 32/1000
sel_vec <- !mean_met$Metric %in% 'Anom'
mean_met$Value[sel_vec] <- 0.032 * mean_met$Value[sel_vec]

# make columns as factors for correct row, column order w/ dcast
mean_met$Input <- factor(mean_met$Input, levels = c('Observed', 'Detided'))
mean_met$Metab <- factor(mean_met$Metab, levels = c('Pg', 'Rt', 'NEM'))
mean_met$Metric <- factor(mean_met$Metric, levels = c('Avg', 'se', 'Anom'))

# dcast long to wide
mean_met <- dcast(mean_met, Site + Input ~ Metab + Metric, value.var = 'Value')

##
# % changes for specific sites

# elk
elk_per <- mean_met[grep('ELK', mean_met$Site), c('Pg_Avg', 'Rt_Avg')]
elk_per <- -100 * (elk_per[1, ] - elk_per[2, ])/elk_per[1, ]

# pdb
pdb_per <- mean_met[grep('PDB', mean_met$Site), c('Pg_Avg', 'Rt_Avg')]
pdb_per <- -100 * (pdb_per[1, ] - pdb_per[2, ])/pdb_per[1, ]
